Wait-free Queues with Polygarithmic Step Complexity

### 2022-05-04 article/psuedocode.tex
infinte loop problem test instance

\newcommand{\sub}[1]{\textsubscript{#1}}
\renewcommand{\tt}[1]{\texttt{#1}}
\renewcommand{\sl}[1]{\textsl{#1}}
\renewcommand{\it}[1]{\textit{#1}}
\renewcommand{\bf}[1]{\textbf{#1}}
\newcommand{\cmt}[1]{\Comment{#1}}

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

@@

%\Statex \bf{Structure}

@@

\Statex $\diamondsuit$ \tt{\sl{Local}}
\begin{itemize}
\item \tt{\sl{*Node} leaf:} \sf{ a pointer to the process's leaf in the tree.}
\end{itemize}

\Statex
\Statex $\diamondsuit$ \tt{\sl{Structures}}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList} blocks}
\item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent. Since it is incremented after propagating, it may be behind by 1.}
\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Root} extends \sl{Node}}
\begin{itemize}

\item \tt{\sl{PBRT} blocks}

  \textsf{\tt{BlockList} is implemented with a persistent red-black tree.}

\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{NonRootNode} extends \sl{Node}}
\begin{itemize}

\item \tt{\sl{Block[]} blocks}

  \textsf{\tt{BlockList} is implemented with an array.}

\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{NonRootNode}}
\begin{itemize}

  \item \tt{\sl{int} \tt{last\sub{done}}}

  \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the. A block is finished if all of its operations are finished. \tt{enqueue(e)} is finished if \tt{e} is returned by some \tt{dequeue()} and \tt{dequeue()} is finished when it computes its response. \it{put the definitions before the pseudocode}}

\end{itemize}
\pagebreak

\Statex $\blacktriangleright$ \tt{\sl{Block}} 
%\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
\cmt{For a \tt{block} in a \tt{blocklist} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}. \it{put the definitions before the pseudocode}}

\begin{itemize}
%  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
%  \textsf{: \# enqueue, dequeue operations in the block}
%  \item \tt{\sl{int} num, sum}
%  \textsf{: total \# operations in block, prefix sum of \tt{num}}

  \item \tt{\sl{int} group}
  \textsf{: the value read from \tt{num\sub{propagated}} when appending this block to the node.}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
\begin{itemize}
  \item \tt{\sl{Object} element}
  \textsf{: Each block in a leaf represents a single operation. For \tt{enqueue} operations element is the input of the \tt{enqueue} and for \tt{dequeue} operations it is \tt{null}.}

    \item \tt{\sl{Object} response}
  \textsf{: stores the response of the operation in the \tt{LeafBlock}.}

    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{: \# enqueue, dequeue operations in the prefix for the block}

\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
\begin{itemize}
    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{:~~index of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}
  \textsf{: \# enqueue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}
  \textsf{: \# dequeue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}
  \textsf{: \# enqueue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}
  \textsf{: \# dequeue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
\begin{itemize}
  \item \tt{\sl{int} length}
  \textsf{: length of the queue after performing all operations in the prefix for this block}
%  \item \tt{\sl{int} sum\sub{non-null deq}}
%  \textsf{: count of non-null dequeus up to this block}
  \item \tt{\sl{counter} num\sub{finished}}
  \textsf{: number of finished operations in the block}
    \item \tt{\sl{int} order}
  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
\end{itemize}

%\Statex $\blacktriangleright$ \tt{Conventions}
%\begin{itemize}
%  \item \tt{i} : index of ith operation in the tree
%  \item \tt{j} : index of jth operation in a node
%  \item \tt{b\sub{n}} : index of the block containing the operation n based on the scope
%  \item Also we are not going to refer to blocks directly, only with their indices. Except while constructing a new block.
%\end{itemize}

@@

\begin{footnotesize}

\it{Variable naming:}
\begin{itemize}
  \item \tt{b\sub{op}}: index of the block containing operation\tt{op}
  \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
\end{itemize}

\it{Abbreviations:}
\begin{itemize}
 \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
 \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\}, blocks[0].num\sub{x}=0})
\end{itemize}
\end{footnotesize}

%##########################################

\begin{algorithm}
\caption{\tt{\sl{Queue}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{200}
\begin{multicols}{2}

\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
\State \tt{newBlock.element= e}
%\State \tt{b.num\sub{enq}=1}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}+1}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}}
\State \tt{leaf.}\Call{Append}{newBlock}
\EndFunction{Enqueue}

\Statex
\Function{Object}{Dequeue()}{}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \Return\tt{leaf.}\Call{HelpDequeue()}{}
\EndFunction{Dequeue}

\Function{int, int}{FindResponse}{int i, int b}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
\State \Return \tt{<-1,-->}
\Else
\Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}. $r$th non-null dequeue returns the element of th $r$th enqueue. We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= length - \#enqueues. Note that the $i$th dequeue in the given block is not a non-null dequeue.}
\State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
\State \Return \tt{<root.blocks.get(enq, r\sub{enq}).order, r\sub{enq}>}
\EndIf
\EndFunction{FindResponse}

%##########################################

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{300}
\begin{multicols}{2}

\Function{void}{Propagate()}{}
\If{\bf{not} \Call{Refresh()}{}}
\State \Call{Refresh()}{} \cmt{Lemma Double Refresh}
\EndIf
\If{\tt{this} \bf{is not} \tt{root}}
\State \tt{parent.}\Call{Propagate()}{}
\EndIf
\EndFunction{Propagate}

\Function{boolean}{Refresh()}{}
\State \tt{s= size}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, s)}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{size, h, h+1}}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
\Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}

\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}. Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
%\State \Return \tt{result block's index}
\EndFunction{BSearch}

\pagebreak

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}
\State \tt{newBlock.order= i}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.size} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{this\sub{dir}= dir.num\sub{propagated}}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}
%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
\EndIf

\State \Return \tt{<b, np\sub{left}, np\sub{right}>}
\EndFunction{CreateBlock}

%##########################################

\begin{algorithm}
\caption{Node}
\begin{algorithmic}[1]
\setcounter{ALG@line}{400}

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element}
\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \cmt{Search range of left child's subblocks of \tt{blocks[b]}.}
\State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.}
\State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
\EndIf
\EndFunction{GetEnq}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the rank of $i$th dequeue in the $b$th block of the node, among the dequeues in the root.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
\If{\tt{dir {\keywordfont is} right}}
%\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
\State \tt{i+= blocks[superBlock].sum\sub{deq-left}} \cmt{consider the dequeues from the right child}
\EndIf
\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
\EndIf
\EndFunction{Index}

\end{algorithmic}
\end{algorithm}

%##########################################

\begin{algorithm}
\caption{Leaf}
\begin{algorithmic}[1]
\setcounter{ALG@line}{500}

\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
\State \tt{size+=1} \label{appendEnd} 
\State \tt{blk.group= size} \label{appendStart} 
\State \tt{blocks[size]= blk} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

>>>>

\newcommand{\sub}[1]{\textsubscript{#1}}
\renewcommand{\tt}[1]{\tt{#1}}
\renewcommand{\sl}[1]{\textsl{#1}}
\renewcommand{\it}[1]{\textit{#1}}
\renewcommand{\bf}[1]{\textbf{#1}}
\newcommand{\nf}[1]{\normalfont{#1}}
\newcommand{\cmt}[1]{\Comment{#1}}

%\usepackage{amsmath,amssymb,amsthm}
%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}[theorem]{Lemma}
%\theoremstyle{definition}
%\newtheorem{definition}[theorem]{Definition}

@@

\Statex \bf{Structure}

@@

% \Statex $\diamondsuit$ \tt{\sl{Local}}
% \begin{itemize}
% \item \tt{\sl{*Node} leaf:} \sf{ a pointer to the process's leaf in the tree.}
% \end{itemize}

% \Statex
% \Statex $\diamondsuit$ \tt{\sl{Structures}}

% \Statex $\blacktriangleright$ \tt{\sl{Node}}
% \begin{itemize}
% \item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
% \item \tt{\sl{BlockList} blocks}
%   \textsf{\tt{BlockList} is implemented with an array.}
% \item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
% \item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent. Since it is incremented after propagating, it may be behind by 1.}
% \item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
% \end{itemize}

% %\Statex $\blacktriangleright$ \tt{\sl{Root} extends \sl{Node}}
% %\begin{itemize}
% %  
% %\item \tt{\sl{PBRT} blocks}
% %
% %  \textsf{\tt{BlockList} is implemented with a persistent red-black tree.}
% %  
% %\end{itemize}

% %\Statex $\blacktriangleright$ \tt{\sl{NonRootNode} extends \sl{Node}}
% %\begin{itemize}

% %\end{itemize}

% \Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{NonRootNode}}
% \begin{itemize}

%   \item \tt{\sl{int} \tt{last\sub{done}}}

%   \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the. A block is finished if all of its operations are finished. \tt{enqueue(e)} is finished if \tt{e} is returned by some \tt{dequeue()} and \tt{dequeue()} is finished when it computes its response. \it{put the definitions before the pseudocode}}

% \end{itemize}
% \pagebreak

% \Statex $\blacktriangleright$ \tt{\sl{Block}} 
% %\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
% \cmt{For a \tt{block} in a \tt{blocklist} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}. \it{put the definitions before the pseudocode}}

% \begin{itemize}
% %  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
% %  \textsf{: \# enqueue, dequeue operations in the block}
% %  \item \tt{\sl{int} num, sum}
% %  \textsf{: total \# operations in block, prefix sum of \tt{num}}

%   \item \tt{\sl{int} group}
%   \textsf{: the value read from \tt{num\sub{propagated}} when appending this block to the node.}
% \end{itemize}

% \Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
% \begin{itemize}
%   \item \tt{\sl{Object} element}
%   \textsf{: Each block in a leaf represents a single operation. For \tt{enqueue} operations element is the input of the \tt{enqueue} and for \tt{dequeue} operations it is \tt{null}.}

%     \item \tt{\sl{Object} response}
%   \textsf{: stores the response of the operation in the \tt{LeafBlock}.}

%     \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
%   \textsf{: \# enqueue, dequeue operations in the prefix for the block}

% \end{itemize}

% \Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
% \begin{itemize}
%     \item \tt{\sl{int} end\sub{left}, end\sub{right}}
%   \textsf{:~~index of the last subblock of the block in the left and right child}
%   \item \tt{\sl{int} sum\sub{enq-left}}
%   \textsf{: \# enqueue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
%   \item \tt{\sl{int} sum\sub{deq-left}}
%   \textsf{: \# dequeue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
%   \item \tt{\sl{int} sum\sub{enq-right}}
%   \textsf{: \# enqueue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
%   \item \tt{\sl{int} sum\sub{deq-right}}
%   \textsf{: \# dequeue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
% \end{itemize}

% \Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
% \begin{itemize}
%   \item \tt{\sl{int} length}
%   \textsf{: length of the queue after performing all operations in the prefix for this block}
% %  \item \tt{\sl{int} sum\sub{non-null deq}}
% %  \textsf{: count of non-null dequeus up to this block}
%   \item \tt{\sl{counter} num\sub{finished}}
%   \textsf{: number of finished operations in the block}
%     \item \tt{\sl{int} order}
%   \textsf{: the index of the block in the \tt{BlockList} containing the block.}
% \end{itemize}

% %\Statex $\blacktriangleright$ \tt{Conventions}
% %\begin{itemize}
% %  \item \tt{i} : index of ith operation in the tree
% %  \item \tt{j} : index of jth operation in a node
% %  \item \tt{b\sub{n}} : index of the block containing the operation n based on the scope
% %  \item Also we are not going to refer to blocks directly, only with their indices. Except while constructing a new block.
% %\end{itemize}

@@

% \begin{footnotesize}

% \it{Variable naming:}
% \begin{itemize}
%   \item \tt{b\sub{op}}: index of the block containing operation\tt{op}
%   \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
% \end{itemize}

% \it{Abbreviations:}
% \begin{itemize}
%  \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
%   \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\}, blocks[0].num\sub{x}=0})
% \end{itemize}
% \end{footnotesize}

% %##########################################

% \begin{algorithm}
% \caption{\tt{\sl{Queue}}}
% \begin{algorithmic}[1]
% \setcounter{ALG@line}{200}
% \begin{multicols}{2}

% \Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
% \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
% \State \tt{newBlock.element= e}
% %\State \tt{b.num\sub{enq}=1}
% \State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}+1}
% \State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}}
% \State \tt{leaf.}\Call{Append}{newBlock}
% \EndFunction{Enqueue}

% \Statex
% \Function{Object}{Dequeue()}{}
% \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
% \State \tt{newBlock.element= null}
% \State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
% \State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
% \State \tt{leaf.}\Call{Append}{newBlock}
% \State \Return\tt{leaf.}\Call{HelpDequeue()}{}
% \EndFunction{Dequeue}

% \Statex
% \pagebreak

% \Function{int, int}{FindResponse}{int i, int b}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns \tt{<-1,-->} if the queue is empty.}
% \If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
% \State \Return \tt{<-1,-->}
% \Else
% \Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}. $r$th non-null dequeue returns the element of th $r$th enqueue. We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= length - \#enqueues. Note that the $i$th dequeue in the given block is not a non-null dequeue.}
% \State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
% \State \Return \tt{<root.blocks.get(enq, r\sub{enq}).order, r\sub{enq}>}
% \EndIf
% \EndFunction{FindResponse}

% \end{multicols}
% \end{algorithmic}
% \end{algorithm}

% %##########################################

% \begin{algorithm}
% \caption{\tt{\sl{Node}}}
% \begin{algorithmic}[1]
% \setcounter{ALG@line}{300}
% \begin{multicols}{2}

% \Function{void}{Propagate()}{}
% \If{\bf{not} \Call{Refresh()}{}}
% \State \Call{Refresh()}{} \cmt{Lemma Double Refresh}
% \EndIf
% \If{\tt{this} \bf{is not} \tt{root}}
% \State \tt{parent.}\Call{Propagate()}{}
% \EndIf
% \EndFunction{Propagate}

% \Statex

% \Function{boolean}{Refresh()}{}
% \State \tt{s= size}
% %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
% \State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
% \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
% \ElsIf{\tt{blocks.tryAppend(new, s)}}
% \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
% \State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
% \State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
% \EndFor
% \State \tt{\Call{CAS}{size, h, h+1}}
% \State \Return{\tt{true}}
% \Else
% \State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.}
% \State \Return{ \tt{false}}
% \EndIf
% \EndFunction{Refresh}

% \Statex

% \Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
% \Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}

% \Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}. Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
% %\State \Return \tt{result block's index}
% \EndFunction{BSearch}

% \pagebreak

% \Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
% \Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
% \State \tt{block newBlock= \Call{new}{\sl{block}}}
% \State \tt{newBlock.group= num\sub{propagated}}
% \State \tt{newBlock.order= i}
% \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
% \State \tt{index\sub{last}= dir.size} \label{lastLine}
% \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
% \State \tt{newBlock.end\sub{dir}= index\sub{last}}
% \State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
% \State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
% \State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
% \State \tt{this\sub{dir}= dir.num\sub{propagated}}
% \State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
% \State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
% \EndFor
% %\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
% %\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
% %\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
% %\State \tt{b.sum= n.blocks[i-1].sum + b.num}

% \If{\tt{this} \bf{is} \tt{root}}
% \State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}
% %\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
% \EndIf

% \State \Return \tt{<b, np\sub{left}, np\sub{right}>}
% \EndFunction{CreateBlock}

% \end{multicols}
% \end{algorithmic}
% \end{algorithm}

% %##########################################

% \begin{algorithm}
% \caption{Node}
% \begin{algorithmic}[1]
% \setcounter{ALG@line}{400}

% \Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
% \Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
% \If{\tt{this} \bf{is} \tt{leaf}}
% \State\Return \tt{blocks[b].element}
% \ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
% \State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \cmt{Search range of left child's subblocks of \tt{blocks[b]}.}
% \State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
% \Else
% \State \tt{i= i-blocks[b].num\sub{enq-left}}
% \State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.}
% \State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
% \EndIf
% \EndFunction{GetEnq}

% \Statex
% \Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
% \Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the rank of $i$th dequeue in the $b$th block of the node, among the dequeues in the root.}
% \If{\tt{this} \bf{is} \tt{root}}
% \State\Return \tt{<b, i>}
% \Else
% \State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
% \State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
% \Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
% \If{\tt{dir {\keywordfont is} right}}
% %\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
% \State \tt{i+= blocks[superBlock].sum\sub{deq-left}} \cmt{consider the dequeues from the right child}
% \EndIf
% \State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
% \EndIf
% \EndFunction{Index}

% \end{algorithmic}
% \end{algorithm}

% %##########################################

% \begin{algorithm}
% \caption{Leaf}
% \begin{algorithmic}[1]
% \setcounter{ALG@line}{500}

% \Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
% \State \tt{size+=1} \label{appendEnd} 
% \State \tt{blk.group= size} \label{appendStart} 
% \State \tt{blocks[size]= blk} 
% \State \tt{parent.}\Call{Propagate()}{} 
% \EndFunction{Append}

% \Statex
